Experience Report: Haskell in Computational Biology

### 2012-03-02 mrfy_experience_report.tex
clean up types, rewrite intro, tweak diagrams, add alignment diagram

As a brief refresher in the standard dogma of genetics, recall that genes 
(strands of deoxyribonucleic acid (DNA), which are sequences of the nucleotides 
adenosine, cytosine, guanine, and thymine, represented by the letters A, C, G, 
and T) are transcribed into ribonucleic acid (RNA). Some RNA is then translated 
into the 20 naturally-occuring amino acids, which form peptide chains -- 
proteins -- which fold into complex structures in the cell. These proteins are 
cellular machinery, performing the varied functions an organism needs to 
survive.

However, determining what newly sequenced genes actually do -- when they become 
proteins, what structures those proteins fold into, and what functions they 
perform -- has not kept pace. Determining the atomic coordinates of a newly 
discovered protein may require as much as two graduate-student-years in a lab! 
Fortunately, determining the full structure of a protein is not always 
necessary. Often, we can identify well-known proteins of similar sequence or 
structure to a new protein, and thus make educated guesses as to the structure 
and function of the new protein.

The problem of taking a newly found protein sequence about which only the 
sequence is known, and determining what existing proteins of known structure 
and function most closely resemble that new protein, is known as homology 
detection. When proteins of similar sequence can be found, this problem is 
largely solved. To solve the homology detection problem, computational 
biologists have developed fast, approximate methods for determining the 
structure and function of proteins based on their sequence. The most popular 
software for homology detection is called HMMER, which uses a hidden Markov 
model to capture evolutionary change.

However, when there are no known proteins of similar sequence, only 
evolutionarily distant proteins that may share similar structure, this problem 
is called remote homology detection. Hidden Markov model approaches begin to 
fail as sequence similarity falls off. Recently, Markov random field 
approaches, which capture non-local interactions in the protein structure, have 
been shown to perform well. However, these approaches face the challenge of 
increased computational complexity; the SMURF program exhibits exponential time 
complexity, while SMURFLite (work by one of the authors) bounds the exponent by 
simplifying the Markov random field's dependency graph. MRFy is a 
stochastic search approach to Markov random fields.

SMURF, SMURFLite, and MRFy all attempt to compute the optimal parse, or 
alignment, of a protein sequence onto the directed graph representing a Markov 
random field (see Figure \ref{mrf}). A parse is a mapping of the letters of a 
query sequence onto the states of a markov Random field, while an optimal parse 
is the parse exhibiting the greatest probability once. The algorithm underlying 
MRFy grew out of an understanding of the limitations of the SMURF (Structural 
Motifs Using Random Fields) algorithm, which used multidimensional dynamic 
programming to compute the optimal parse of a query sequence onto a Markov 
random field. Rather than attempt to compute exactly the optimal parse, MRFy 
treats the non-local interactions (which correspond to hydrogen-bonded ``beta 
strands'' in the protein) as ``beads'' which can be slid along a string. In 
this analogy, the string is the sequence of amino acids of the query protein. A 
given placement of these beads corresponds to an assignment of particular amino 
acid residues being hydrogen bonded to other particular amino acid residues, 
and thus MRFy computes a probability based on observed frequencies of residues 
hydrogen bonding in beta strands~\cite{betawrappro}. In between these beta 
strands, the Markov random field is in fact a traditional hidden Markov model, 
whose optimal solution can be easily computed via dynamic programming (known as 
the Viterbi algorithm). MRFy performs a stochastic search to place these beta 
strands, based on an objective function that combines the probability computed 
by the Viterbi algorithm with that computed from the paired, hydrogen-bonded 
residues in the beta strands.

\begin{figure} 
\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}} 
\caption{The directed graph representing the states of a MRFy Markov random 
field finite state automaton. Shaded nodes represent beta-strand positions, 
while the dashed edges represent the long-range dependencies due to hydrogen 
bonding.}\label{mrf} \end{figure}

@@

\subsection{Beautiful code}
MRFy contains two distinct algorithms: the classical Viterbi 
dynamic-programming algorithm for hidden Markov models, and a stochastic search 
algorithm for placing beta-strand positions. This stochastic search algorithm 
implements several search heuristics: random-mutation hill climbing, simulated 
annealing, multistart simulated annealing, and a genetic algorithm.

@@

@@

\begin{figure} \centerline{\includegraphics[width=8cm]{Plan7.pdf}} \caption{The 
directed graph representing the states of a traditional HMMER ``Plan7'' hidden 
Markov model. Only 7 of the possible 9 state transitions are allowed, which 
simplifies the dynamic programming. Insertion states represent mutations that 
insert an amino acid, deletion states represent mutations that lose an amino 
acid, and match states represent either conserved amino acids (those with no 
changes) or mutations that substitute one amino acid for another. Each match or 
insertion state has an associated table of emission probabilities (here 
illustrated only for the first two states)}\label{plan7} \end{figure}

>>>>

Proteins are cellular machinery. They interact with one another and with other 
molecules to carry out the functions of living cells: metabolism, regulation, 
signalling, and more. The function of a protein is determined by its structure, 
and this structure in turn is determined by the specific sequence of the twenty 
different amino acids that make up the protein, but proteins with significantly 
different amino acid sequences may share similar structure and function. This 
amino acid sequence is ultimately determined by the DNA sequence we call a 
gene. Given the genetic sequence of a protein, biologists wish to know its 
specific cellular function. The best path to this is to find other proteins of 
similar structure, which likely share similar function. This problem is called 
\textit{homology detection}.

Computational biologists address the homology detection problem by building 
algorithms to find similar proteins given an amino acid sequence. When there 
are nearly-identical sequences for proteins of known structure, this is a 
simple string-matching problem. When there are no identical sequences, but 
still some reasonably similar sequences, one family of algorithms called 
\textit{hidden Markov models} excels at homology detection. We will explain an 
algorithm for detecting reasonably similar sequences, and then move on to 
explain MRFy, our novel approach for detecting more distantly related sequences 
for proteins that share similar structure and function.

% As a brief refresher in the standard dogma of genetics, recall that genes 
% (strands of deoxyribonucleic acid (DNA), which are sequences of the nucleotides 
% adenosine, cytosine, guanine, and thymine, represented by the letters A, C, G, 
% and T) are transcribed into ribonucleic acid (RNA). Some RNA is then translated 
% into the 20 naturally-occuring amino acids, which form peptide chains -- 
% proteins -- which fold into complex structures in the cell. These proteins are 
% cellular machinery, performing the varied functions an organism needs to 
% survive.

% However, determining what newly sequenced genes actually do -- when they become 
% proteins, what structures those proteins fold into, and what functions they 
% perform -- has not kept pace. Determining the atomic coordinates of a newly 
% discovered protein may require as much as two graduate-student-years in a lab! 
% Fortunately, determining the full structure of a protein is not always 
% necessary. Often, we can identify well-known proteins of similar sequence or 
% structure to a new protein, and thus make educated guesses as to the structure 
% and function of the new protein.
% 
% The problem of taking a newly found protein sequence about which only the 
% sequence is known, and determining what existing proteins of known structure 
% and function most closely resemble that new protein, is known as homology 
% detection. When proteins of similar sequence can be found, this problem is 
% largely solved. To solve the homology detection problem, computational 
% biologists have developed fast, approximate methods for determining the 
% structure and function of proteins based on their sequence. The most popular 
% software for homology detection is called HMMER, which uses a hidden Markov 
% model to capture evolutionary change.

% However, when there are no known proteins of similar sequence, only 
% evolutionarily distant proteins that may share similar structure, this problem 
% is called remote homology detection. Hidden Markov model approaches begin to 
% fail as sequence similarity falls off. Recently, Markov random field 
% approaches, which capture non-local interactions in the protein structure, have 
% been shown to perform well. However, these approaches face the challenge of 
% increased computational complexity; the SMURF program exhibits exponential time 
% complexity, while SMURFLite (work by one of the authors) bounds the exponent by 
% simplifying the Markov random field's dependency graph. MRFy is a 
% stochastic search approach to Markov random fields.

@@

\subsection{The algorithms}

A biologist may use software such as HMMER or MRFy to test a hypothesis about 
the function of a single, newly discovered protein, or she may use the software 
to scan an entire newly sequenced genome against a library of known protein 
structures. This latter case demands particularly high-performance software.

%% move into later section

%% add something about groups of related proteins aligned (put into correspondence)

SMURF, SMURFLite, and MRFy all attempt to compute the optimal parse, or 
alignment, of a protein sequence onto the directed graph representing a Markov 
random field (see Figure \ref{mrf}). A parse is a mapping of the letters of a 
query sequence onto the states of a markov Random field (and the corresponding path through the finite state automaton), while an optimal parse 
is the parse exhibiting the greatest probability. The algorithm underlying 
MRFy grew out of an understanding of the limitations of the SMURF (Structural 
Motifs Using Random Fields) algorithm, which used multidimensional dynamic 
programming to compute the optimal parse of a query sequence onto a Markov 
random field. Rather than attempt to compute exactly the optimal parse, MRFy 
treats the non-local interactions (which correspond to hydrogen-bonded ``beta 
strands'' in the protein) as ``beads'' which can be slid along a string. In 
this analogy, the string is the sequence of amino acids of the query protein. A 
given placement of these beads corresponds to an assignment of particular amino 
acid residues being hydrogen bonded to other particular amino acid residues, 
and thus MRFy computes a probability based on observed frequencies of residues 
hydrogen bonding in beta strands~\cite{betawrappro}. In between these beta 
strands, the Markov random field is in fact a traditional hidden Markov model, 
whose optimal solution can be easily computed via dynamic programming (known as 
the Viterbi algorithm). MRFy performs a stochastic search to place these beta 
strands, based on an objective function that combines the probability computed 
by the Viterbi algorithm with that computed from the paired, hydrogen-bonded 
residues in the beta strands.

%% "nailing down" certain subsequences of the model; between beta strands we have HMMs

\begin{figure} 
\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}} 
\caption{The directed graph representing the states of a MRFy Markov random 
field finite state automaton. Shaded nodes represent beta-strand positions which have been fixed into match states, 
while the dashed edges represent the long-range dependencies due to hydrogen 
bonding. Note that in between the beta-strand regions are normal HMMs.}\label{mrf} \end{figure}

MRFy contains two distinct algorithms: the classical Viterbi 
dynamic-programming algorithm for hidden Markov models, and a stochastic search 
algorithm for placing beta-strand positions. This stochastic search algorithm 
implements several search heuristics: random-mutation hill climbing, simulated 
annealing, multistart simulated annealing, and a genetic algorithm.

@@

%% break the viterbi types up and insert into the Plan7 explanation.

@@

%% group nodes, illustrate an alignment

\begin{figure}\centerline{\includegraphics[width=6cm]{alignment.pdf}} \caption{Protein sequences are put into alignment based on structural or sequence similarities. Each column of the alignment corresponds to a node of the hidden Markov model; columns with many gaps will induce highly probable insertion states in the HMM, while columns with fewer gaps will induce more-probable deletion and match states. Columns with strong consensus towards a particular residue will induce match states with high emission probability for that residue.}\label{alignment} \end{figure}

%% how do proteins change due to evolution?
%% explain insertion, deletion, substitution.

%% what looks like an intimidating geometry is completely explained by the fact
%% that we have insertions, deletions, and substitutions.

\begin{figure} \centerline{\includegraphics[width=8cm]{Plan7.pdf}} \caption{The 
directed graph representing the states of a traditional HMMER ``Plan7'' hidden 
Markov model. Only 7 of the possible 9 state transitions are allowed, which 
simplifies the dynamic programming. Insertion states represent mutations that 
insert an amino acid, deletion states represent mutations that lose an amino 
acid, and match states represent either conserved amino acids (those with no 
changes) or mutations that substitute one amino acid for another. Each match or 
insertion state has an associated table of emission probabilities (here 
illustrated only for the first two states). Triplets of insertion, deletion, and match states are grouped into ``nodes,'' represented here with dashed ovals; each node corresponds to a column in the original alignment used to train the HMM.}\label{plan7} \end{figure}
